---
title: "Lab3"
author: "Michal Malyska"
date: "January 22 2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Overview

Today we are looking at data on infant deaths (deaths in the first year of life) in the US. The dataset `infant` contains information on all deaths to the 2012 birth cohort. For today, we are interested in investigating differences in neonatal deaths (i.e. deaths in the first month of life) and cause of death. 

## What to hand in

As with last week, please push your Rmd and compiled document (html or pdf) to GitHub. **The questions for this week are dispersed throughout the lab.**

## The dataset

Read it in and have a look to see what's in there. Variables are

- `sex`: sex of baby
- `aged`: age at death (in days)
- `race`: race of mother
- `gest`: gestation in weeks
- `ucod`: cause of death (ICD-10 code)
- `cod`: cause of death, descriptive groups
- `mom_age`: mother age in years
- `mom_age_group`: mother age group

```{r library load}
library(tidyverse)
library(here)
library(kableExtra)
library(GGally)
d <- read_rds(here("data", "infant.RDS"))
head(d)
```

# Descriptives

Let's create some new variables that will be useful:

- `neo_death`: equals 1 if the death occurred in the first 28 days
- `preterm`: equals 1 if gestational age is less than 37 weeks
- `cod_group`: reduced number of categories of cause of death

Also, removing the observations where we don't know gestational age or the mother's age.

```{r create new vars}
d <- d %>%
	mutate(
		neo_death = ifelse(aged <= 28, 1, 0),
		cod_group = case_when(
			str_starts(cod, "peri") ~ "perinatal",
			cod %in% c("other", "unknown") ~ "oth_unk",
			cod %in% c("sids", "maltreatment", "infection") ~ "exogenous",
			cod %in% c("resp", "heart") ~ "resp_heart",
			TRUE ~ cod
		),
		preterm = ifelse(gest < 37, 1, 0)
	) %>%
	filter(gest < 99,!is.na(mom_age_group))
```

## Distribution of gestational ages

Let's plot the distribution of gestational ages by race. It's quite bi-modal. Notice the difference in densities by race. 

```{r histogram of ages}
d %>% ggplot(aes(gest, fill = race)) +
	geom_histogram(position = 'dodge', aes(y = ..density..)) +
	theme_minimal()
```

## Question 1

Calculate the proportion of deaths that are neonatal by race and prematurity. Which group has the highest proportion of neonatal deaths?

```{r Proportion of deaths}
q1_df <- d %>% group_by(race, preterm) %>%
	summarize(prop_neonatal = mean(neo_death))

kableExtra::kable(q1_df)
```

Highest proportion of neonatal deaths are in the race "NHW" (non-hispanic white) and for the
premature births. In general it seems that the premature births have an 
incredibly higher neonatal death rate. 

## Causes of death

Let's make `cod_group` a factor with congenital malformations as the reference.

```{r add cod_group}
d <- d %>% 
  mutate(cod_group = factor(cod_group, levels = c("cong_mal", "perinatal", "resp_heart", "exogenous", "oth_unk"),
                         labels = c("cong_mal", "perinatal", "resp_heart", "exogenous", "oth_unk")))
```

The following code calculates the proportion of deaths by cause group, race, sex and prematurity

```{r df of prop causes}
prop_cause <- d %>%
	group_by(race, preterm, sex, cod_group) %>%
	summarise(n = n()) %>%
	group_by(race, preterm, sex) %>%
	mutate(prop = n / sum(n)) %>%
	ungroup() %>%
	mutate(preterm = ifelse(preterm == 1, "pre-term", "full-term"))
```


## Question 2

Using the `prop_cause` above, filter to just look at female babies, and make a graph to help visualize differences in cause by race and prematurity. 

```{r graph on race and prematurity for female babies}

prop_cause %>% filter(sex == "F") %>%
	ggplot(aes(
		x = race,
		facet = preterm,
		y = prop,
		color = cod_group )) +
	geom_point() +
	facet_wrap(. ~ preterm)

```


# Logistic regression

First, let's do logistic regression to explore differences in neonatal deaths. Here's a  model with prematurity, sex, race, and mom's age

```{r log reg neo_deat}
mod <- glm(neo_death~ preterm + sex + race + race:preterm + mom_age, data = d, family = binomial)
summary(mod)
```


## Question 3

Rerun the model above with instead of `mom_age`, include a new variable `mom_age_c` which centers mother's age around its mean. 

```{r log_reg mom_age normalized}

d_model <- d %>% mutate(mom_age_c = scale(mom_age, scale = FALSE))

mod_age_norm <- glm(neo_death~ preterm + sex + race + race:preterm + mom_age_c, data = d_model, family = binomial)
summary(mod_age_norm)


```


## Question 4

```{r extract coefficients}

GGally::ggcoef(mod_age_norm,exponentiate = TRUE) +
	labs(title = "Coefficients of logistic regression")

coefs <- mod_age_norm$coefficients

```


Interpret the `preterm`, `race` and the interaction `preterm:race` coefficients. 

The preterm coefficient with a value of `r coefs["preterm"]` which means that 
babies born prematurely are `r exp(coefs["preterm"])` time more likely to have a
neonatal death.

The race coefficient with a value of `r coefs["raceNHW"]` which means that 
babies born as to non hispanic whites are `r exp(coefs["raceNHW"])` time more likely to have a
neonatal death.

The preterm:raceNHW coefficient with a value of `r coefs["preterm:raceNHW"]` which means that 
babies born preterm to non hispanic whites are a further  `r exp(coefs["preterm:raceNHW"])` time more likely to have a neonatal death on top of the other coefficients (for race:NHW and preterm).

# Multinomial regression

Now let's do multinomial regression with cause of death as the outcome. We need to get the data in a different format to run the regression:

```{r multinomial regression data prep}

d$mom_age_c <- d$mom_age - mean(d$mom_age)
d_wide <- d %>%
	group_by(sex, race, cod_group, preterm, mom_age_c) %>%
	summarise(deaths = n()) %>%
	pivot_wider(names_from = cod_group, values_from = deaths) %>%
	mutate_all(.funs = funs(ifelse(is.na(.), 0, .)))

d_wide$Y <-	as.matrix(d_wide[, c("cong_mal", "perinatal", "resp_heart", "exogenous", "oth_unk")])

```

Now run the regression

```{r multinomial reressions}

library(nnet)
mod2 <- multinom(Y ~ sex + race + mom_age_c + preterm, data = d_wide)
summary(mod2)

```

## Question 5

Using the `predict` function, find the predicted probabilities of each cause by race, sex and prematurity for the mothers of mean age. You can use this prediction dataframe to get all the combinations you need.

```{r predictions from multinomial regression}

pred_df <- tibble(
	preterm = c(rep(0, 4), rep(1, 4)),
	sex = rep(c(rep("F", 2), rep("M", 2)), 2),
	race = rep(c("NHB", "NHW"), 4),
	mom_age_c = 0)

predictions <- as_tibble(predict(mod2, type = "probs", newdata = pred_df))


pred_df <- cbind(pred_df, predictions) %>%
	pivot_longer(cols = c(names(predictions)), names_to = "cod_group", values_to = "predictions")


```

## Question 6

Plot the predicted probabilities for female babies. 

```{r prediction for female babies}

pred_df_plot <- pred_df %>%
	mutate(preterm = as_factor(if_else(preterm == 0, "Full Term", "Pre-Term")))

pred_df_plot %>% filter(sex == "F") %>%
	ggplot(aes(
		x = race,
		y = predictions,
		facet = preterm,
		color = cod_group
	)) +
	geom_point() +
	facet_wrap(. ~ preterm)

```


## Question 7

What race/prematurity/ cause group has the highest probability? How does this compare to the observed proportion in the same group?

The group with the highest probabilty is the Pre-term babies birth by Non-Hispanic-Black Women
that are born female that have perinatal congenital malformations.

```{r misc calcs, include=FALSE}

observed_prop <- d %>% group_by(race, sex, preterm, cod_group) %>%
	summarize(mean_prop = mean(neo_death)) %>%
	filter(race == "NHB", sex == "F", preterm == 1, cod_group == "perinatal") %>%
	pull(mean_prop)

predicted_prop <- pred_df %>%
	filter(race == "NHB", sex == "F", preterm == 1, cod_group == "perinatal") %>%
	pull(predictions)

```


Observed proportion for this group is `r observed_prop` while the predicted proportion is `r predicted_prop`
which is significantly lower. This is probably not very good. 