---
title: "Assignment 2"
author: "Michal Malyska"
date: "10/02/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(here)
```

```{r data load}
df1 <- read_rds(path = paste0(here(), "/data/ON_mortality.RDS")) %>%
	mutate(age = as.numeric(if_else(age == "110+", "110", age)))
```

# Question 1

$$
\lambda(t) = \alpha e^{\beta t}
$$

## a

$$
S(t) = \exp(-\frac{\alpha}{\beta} (e^{\beta t} - 1))
$$

Showing $S(t)$ (kind of reverse showing since I show that this $S(t)$ implies our
hazard function but still valid.)

$$
\begin{aligned}
\lambda(t) &= - \frac{d}{dt} log(S(t)) \\
&= - \frac{d}{dt} \left( -\frac{\alpha}{\beta} (e^{\beta t} - 1) \right) \\
&= \frac{d}{dt} \left( \frac{\alpha}{\beta} (e^{\beta t} - 1) \right) \\
&= \frac{d}{dt} \left( \frac{\alpha}{\beta} e^{\beta t} \right) \\
&= \alpha e^{\beta t}
\end{aligned}
$$
$$
f(t) = \alpha \exp(\beta t -\frac{\alpha}{\beta} (e^{\beta t} - 1))
$$

Showing $f(t)$

$$
\lambda(t) = \frac{f(t)}{S(t)} \\
f(t) = \lambda(t) S(t)
$$
$$
\begin{aligned}
f(t) &= \lambda(t) S(t) \\
&= \alpha e^{\beta t} * \exp(-\frac{\alpha}{\beta} (e^{\beta t} - 1)) \\
&= \alpha \exp(\beta t-\frac{\alpha}{\beta} (e^{\beta t} - 1))
\end{aligned}
$$

## b

Modal Time of Death (mode of $f(t)$)

$$
\begin{aligned}
\frac{d}{dt} f(t) &= f(t) * (\beta - \alpha e^{\beta t}) = 0 \\
&\implies (\beta - \alpha e^{\beta t}) = 0 \text{ or } f(t) = 0
\end{aligned}
$$

so the mode is at:

$$ 
t = \frac{log(\frac{\beta}{\alpha})}{\beta}
$$

as long as $\alpha < \beta$

otherwise the function is decreasing so:
$$
t = 0
$$

## c

$ h(x) = ae^{bx} = e^{log(a) + bx}$
So:
$log(h(x)) = log(a) + bx$

```{r Fix the data}
df_c <- df1 %>%
	filter(between(age, 40, 100)) %>%
	mutate(loghx = log(hx))

df_1961 <- df_c %>% filter(year == 1961)
df_2011 <- df_c %>% filter(year == 2011)

lm1961 <- lm(loghx ~ age, data = df_1961)
lm2011 <- lm(loghx ~ age, data = df_2011)

coef1961 <- coef(lm1961)
coef2011 <- coef(lm2011)

a1961 = unname(exp(coef1961[1]))
a2011 = unname(exp(coef2011[1]))

b1961 = unname(coef1961[2])
b2011 = unname(coef2011[2])

```

The values for 1961 are: alpha of `r a1961` and beta of `r b1961` compared 
to the values for 2011 : alpha of `r a2011` and beta of `r b2011`

The meaning of alpha is the starting level of mortality (much higher for 1961)
and beta gives the increase in mortality over time which surprisingly is 
higher for 2011. Perhaps lower infant mortality screws with us a tiny bit
and makes it seem like people die faster with age in 2011 than they did in 1961 just
because so many of them already died before 40 where we start.

## d

```{r plotting estimated hazards}

preds_d <- tibble(age = seq(from = 40, to = 100, by = 1))
preds_d$predicted_log_hx_1961 <- predict(lm1961, newdata = preds_d)
preds_d$predicted_log_hx_2011 <- predict(lm2011, newdata = preds_d)
preds_d$actual_log_hx_1961 <- df_1961$loghx
preds_d$actual_log_hx_2011 <- df_2011$loghx

preds_d_long <- preds_d %>% pivot_longer(cols = c("predicted_log_hx_1961", "predicted_log_hx_2011", "actual_log_hx_1961", "actual_log_hx_2011"), names_to = "type", values_to = "log_hazard")
 
preds_d_long %>%
	ggplot(aes(x = age, y = log_hazard, color = type)) +
	geom_point() +
	theme_minimal()

```

They both seem to fit surprisingly well. There are some minor patterns in the
predicted vs actual for 2011 between the ages of 70 and 80 where the 
actual log hazard seems to be lower, and later on when actual log hazard
seems to be higher for those 90+. For 1961 model it seems to be the opposite for
the super old - we overestimate the log hazard for those pushing 100. Overall
I would say the assumption is quite reasonable. 

## e

```{r modal age at death}

mode1961 <- log(b1961 / a1961) / b1961
mode2011 <- log(b2011 / a2011) / b2011

df_e <- df_c %>% filter(year %in% c(2011, 1961))

df_e %>% filter(year == 1961) %>%
	ggplot(aes(x = age, y = dx)) +
	geom_point() +
	geom_vline(xintercept = mode1961, color = "red") +
	theme_minimal() +
	labs(title = "Density for 1961")

df_e %>% filter(year == 2011) %>%
	ggplot(aes(x = age, y = dx)) +
	geom_point() +
	geom_vline(xintercept = mode2011, color = "red") +
	theme_minimal() +
	labs(title = "Density for 2011")

```

## f

I could probably figure this out with map(lm) but I don't wanna right now.

```{r loopin over, echo=TRUE}

years <- unique(df_c$year)
alphas <- rep(NA, length(years))
betas <- rep(NA, length(years))


for (i in 1:length(years)) {
	# print(i)
	 # fit model
	df_model <- df_c %>%
		filter(year == years[i])
	lm_loop <- lm(loghx ~ age, data = df_model)
	coef_model <- coef(lm_loop)

	alphas[i] <- unname(exp(coef_model[1]))
	betas[i] <- unname(coef_model[2])
	
	df_model$loghx_pred <- predict(lm_loop, newdata = df_model)
	
	df_plot <- df_model %>% pivot_longer(cols = c("loghx_pred", "loghx"), names_to = "type", values_to = "log_hazard")
	p <- df_plot %>% ggplot(aes(x = age, y = log_hazard, color = type)) +
	geom_point() +
	theme_minimal() +
	labs(title = paste0("log hazards for predicted vs actual for the year ", years[i]))
	print(p)
}

```

```{r Plot of alphas and betas}

df_all_years <- tibble(year = years, alpha = alphas,beta = betas) %>%
	mutate(mode = log(beta / alpha) / beta)

df_all_years %>%
	ggplot(aes(x = year, y = alpha)) +
	geom_point() +
	theme_minimal() +
	labs(title = "Alphas over time")

df_all_years %>%
	ggplot(aes(x = year, y = beta)) +
	geom_point() +
	theme_minimal() +
	labs(title = "Betas over time")

df_all_years %>%
	ggplot(aes(x = year, y = mode)) +
	geom_point() +
	theme_minimal() +
	labs(title = "Mode over time")

```

Alphas are decreasing over time quite steadily and Betas are increasing over time
almost as steadily. Mode over time increases very steadily.

This means that people now have a lower starting off mortality that increases a
bit faster with age. Overall modal age of death is increasing so people live longer.


# Question 2

```{r Load data}

deaths <- read_rds(here("data", "infant.RDS"))
births <- read_rds(here("data", "births.RDS"))

```


## a

```{r Q2a}

tot_deaths <- deaths %>% group_by(race) %>% summarize(deaths = n())
tot_births <- births %>% group_by(race) %>% summarize(births = sum(births))

summary_tbl <- left_join(tot_deaths, tot_births) %>%
	mutate(IMR = deaths / births)

kableExtra::kable(summary_tbl, format = "html")

mortalities <- summary_tbl %>% pull(IMR)
ratio <- mortalities[1] / mortalities[2]
```

So we have that the ratio of black to white mortality is `r ratio`

## b

```{r Q2b}
library(survival)

deaths_sum <- deaths %>% group_by(race, prematurity, aged) %>% summarize(deaths = n()) 

km_df <- deaths_sum %>% mutate(deaths_tot = cumsum(deaths)) %>% 
	left_join(births) %>%
	mutate(exposure = births - deaths_tot + deaths,
		   hazard = deaths / exposure, 
		   survival = cumprod(1 - hazard),
		   var_comp = hazard / (exposure - deaths),
		   variance  = survival^2 * cumsum(var_comp),
		   survival_plus_two = survival + 2 * sqrt(variance),
		   survival_minus_two = survival - 2 * sqrt(variance)
	)

```

## c

```{r Q2c}
 
# km_df %>% filter(race == "NHB") %>%
# 	ggplot() +
# 	aes(x = aged, y = survival, m = survival_plus_two, n = survival_minus_two,  facet = prematurity) +
# 	geom_path() +
# 	facet_grid(prematurity~., scales = "free_y") +
# 	geom_line(aes(y = survival_plus_two), color = "red") +
# 	geom_line(aes(y = survival_minus_two),color = "red") +
# 	theme_bw() +
# 	labs(title = "NHB")
# 
# km_df %>% filter(race != "NHB") %>%
# 	ggplot() +
# 	aes(x = aged, y = survival, m = survival_plus_two, n = survival_minus_two,  facet = prematurity) +
# 	geom_path() +
# 	facet_grid(prematurity~., scales = "free_y") +
# 	geom_line(aes(y = survival_plus_two), color = "red") +
# 	geom_line(aes(y = survival_minus_two),color = "red") +
# 	theme_bw() +
# 	labs(title = "NHW")
	
km_df %>% 
	ggplot() +
	aes(x = aged, y = survival, m = survival_plus_two, n = survival_minus_two,  facet = prematurity) +
	geom_path() +
	facet_grid(prematurity~race, scales = "free_y") +
	geom_line(aes(y = survival_plus_two), color = "red") +
	geom_line(aes(y = survival_minus_two),color = "red") +
	theme_bw() +
	labs(title = "Kaplan Meier Survival curves by age faceted by race and prematurity of birth")
```


## d

Because there are just so god damn many full-term babies. The curve for NHB for
those is not too much lower than for the NHW babies but there are just so 
many of them that this probably skews the entire statistic.


## e

Because the births are not equal so higher mortality is not the same as high
number of deaths. Therefore since there are more NHW in general the number
of deaths will be much higher. 

## f

```{r survsplit}

pch <- deaths %>% mutate(event = 1,
						 aged = if_else(aged == 0, 1e-20, aged))
cuts <- c(1, 7, 14, 28, 60, 90, 120)
cuts <- cuts - 0.000000000001

pch <- survSplit(Surv(time = pch$aged, event = pch$event)~race + prematurity + 1,data = pch, cut = cuts, zero = 0) %>%
	mutate(interval = factor(tstart), interval_len = tstop - tstart)

E_d <- pch %>% group_by(race, prematurity, interval) %>%
	summarize(E_d = sum(interval_len)) %>% pull(E_d)

pch_agg <- pch %>% ungroup() %>% group_by(race, prematurity, interval) %>%
	summarize(deaths = sum(event),
			  interval_len = max(interval_len)) %>%
	mutate(deaths_sum = cumsum(deaths))

exposures_df <- left_join(pch_agg, births) %>% group_by(race, prematurity) %>%
	summarise(alive_at_end = max(births) - sum(deaths))

pch_agg <- left_join(pch_agg, exposures_df) %>%
	mutate(E_a = alive_at_end * interval_len) %>%
	left_join(births) %>%
	rename(total = births)

pch_agg$E_d <- E_d

pch_agg <- pch_agg %>% 
	mutate(E_k = E_d + E_a) %>%
	rename(D_k = deaths)
```

This was absolutely harrowing

```{r piecewise constant}
# 
# t_start <- c(0, 1, 7, 14, 28, 60, 90, 120, 365)
# lengths <- c(1, 6, 7, 14, 32, 30, 30, 245)
# 
# km_df$interval <- cut(km_df$aged, t_start, right = FALSE)
# 
# km_df <- km_df %>% ungroup() %>%
# 	mutate(interval_lenght = case_when(aged < 1 ~ 1,
# 									   aged < 7 ~ 6,
# 									   aged < 14 ~ 7,
# 									   aged < 28 ~ 14,
# 									   aged < 60 ~ 32,
# 									   aged < 90 ~ 30,
# 									   aged < 120 ~ 30,
# 									   TRUE ~ 244))
# 
# interval_df <- km_df %>% group_by(race, prematurity, interval, interval_lenght, births) %>%
# 	summarize(interval_deaths = sum(deaths))
# 
# exposures_df <- interval_df %>% group_by(race, prematurity) %>%
# 	summarise(alive = max(births) - sum(interval_deaths))
# 
# interval_df <- left_join(interval_df, exposures_df) %>%
# 	mutate(E_a = alive * interval_lenght) %>%
# 	left_join(births) %>%
# 	rename(total = births) %>%
# 	mutate(E_d = total - interval_deaths,
# 		   E_k = E_d + E_a)
# 

```

```{r Poisson GLM}

glm_godplzno <- glm(formula = D_k ~ race + prematurity + interval + race*interval + prematurity*interval - 1,  offset = (log(E_k)), data = pch_agg, family = "poisson")
summary(glm_godplzno)

coef_effects <- lapply(split(exp(coef(glm_godplzno)), names(exp(coef(glm_godplzno)))), unname)

fi_ep_nhb <- coef_effects$raceNHB

fi_ep_nhw <- coef_effects$raceNHW

fi_ft_nhb <- coef_effects$raceNHB * coef_effects$`prematurityfull-term`

f120_ft_nhb <- coef_effects$interval119.999999999999 * coef_effects$`prematurityfull-term` * coef_effects$`prematurityfull-term:interval119.999999999999`

f120_ep_nhb <- coef_effects$interval119.999999999999

f120_ft_nhw <- coef_effects$interval119.999999999999 * coef_effects$`prematurityfull-term` * coef_effects$`prematurityfull-term:interval119.999999999999` * coef_effects$raceNHW *
	coef_effects$`raceNHW:interval119.999999999999`

```

Hazard of dying in first interval for extremely preterm and NHB mother is just `r fi_ep_nhb`

a) Relative hazard is `r fi_ep_nhw/fi_ep_nhb` 

b) Relative hazard is `r fi_ft_nhb/fi_ep_nhb`

c) Relative hazard is `r f120_ft_nhb/f120_ep_nhb`

d) Relative hazard is `r f120_ft_nhw/f120_ft_nhb`

## g

```{r survival_prob}

survival_prob <- function(lambdas,
                     cuts, # start and end times that lambdas refers to, starting at 0 and ending at max
                     ## observation time of interest,
                     ## thus length is one more than length of lambda
                     neval = 1000 # at how many points do you want to evaluate S(t) within each interval?
                     ){
  lengthintervals <- rep((cuts[-1] - cuts[-length(cuts)])/neval, each = neval)
  t_seq <- c(0, cumsum(lengthintervals))
  cumulative_hazard <- cumsum(lengthintervals*rep(lambdas, each  = neval))
  surv_probs <- c(1, exp(-cumulative_hazard))
  return(tibble(time = t_seq, surv = surv_probs ))
}

lambdas <- coef_effects$raceNHB * c(1, coef_effects$interval0.999999999999, coef_effects$interval6.999999999999, coef_effects$interval13.999999999999, 
									coef_effects$interval27.999999999999, 
									coef_effects$interval59.999999999999,
									coef_effects$interval89.999999999999,
									coef_effects$interval119.999999999999)
new_cuts = c(0, cuts, 365)

df_surv <- survival_prob(lambdas, new_cuts)

ggplot(aes(time, surv), data = df_surv) +
	geom_line() +
  	ggtitle("Survival for extremely preterm babies to NHB mothers") + 
  	xlab("years") + ylab("S(t)") +
  	theme_bw(base_size = 14)

km_df %>% filter(race == "NHB" & prematurity == "extremely preterm") %>%
	ggplot() +
	aes(x = aged, y = survival, m = survival_plus_two, n = survival_minus_two,  facet = prematurity) +
	geom_line() +
	theme_bw()

km_df %>%
	filter(race == "NHB" & prematurity == "extremely preterm") %>%
	ggplot() +
	geom_line(aes(x = aged, y = survival), color = "red") +
	geom_line(data = df_surv, aes(x = time, y = surv),  color = "black") +
  	ggtitle("Survival for extremely preterm babies to NHB mothers") + 
  	xlab("days") + ylab("S(t)") +
  	theme_bw(base_size = 14) +
	labs(caption = "Kaplan-Meier in red \nPiecewice Constant Hazards in black")

```

# BONUS

```{r bonus time}

pch_tot <- glm(formula = D_k ~ offset(log(E_k))-1 + interval, data = pch_agg, family = "poisson")
pch_tot2 <- glm(formula = D_k ~ interval-1, data = pch_agg, offset = (log(E_k)), family = "poisson")
summary(pch_tot)
summary(pch_tot2)

ce <- lapply(split(exp(coef(pch_tot2)), names(exp(coef(pch_tot2)))), unname)

lambdas <- exp(coef(pch_tot2))
df_surv <- survival_prob(lambdas, new_cuts)

deaths_sum <- deaths %>% group_by(aged) %>% summarize(deaths = n()) 

tot_births <- sum(births$births)

km_df <- deaths_sum %>% mutate(deaths_tot = cumsum(deaths)) %>% 
	mutate(births = tot_births,
		   exposure = births - deaths_tot + deaths,
		   hazard = deaths / exposure, 
		   survival = cumprod(1 - hazard),
		   var_comp = hazard / (exposure - deaths),
		   variance  = survival^2 * cumsum(var_comp),
		   survival_plus_two = survival + 2 * sqrt(variance),
		   survival_minus_two = survival - 2 * sqrt(variance)
	)



km_df %>%
	ggplot() +
	geom_line(aes(x = aged, y = survival), color = "red") +
	geom_line(data = df_surv, aes(x = time, y = surv),  color = "black") +
  	ggtitle("BONUS - Survival for everyone") + 
  	xlab("days") + ylab("S(t)") +
  	theme_bw(base_size = 14) +
	labs(caption = "Kaplan-Meier in red \nPiecewice Constant Hazards in black")
```

