---
title: "Assignment 2"
author: "Michal Malyska"
date: "10/02/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(here)
```

```{r data load}
df1 <- read_rds(path = paste0(here(), "/data/ON_mortality.RDS")) %>%
	mutate(age = as.numeric(if_else(age == "110+", "110", age)))
```

# Question 1

$$
\lambda(t) = \alpha e^{\beta t}
$$

## a

$$
S(t) = \exp(-\frac{\alpha}{\beta} (e^{\beta t} - 1))
$$

Showing $S(t)$ (kind of reverse showing since I show that this $S(t)$ implies our
hazard function but still valid.)

$$
\begin{aligned}
\lambda(t) &= - \frac{d}{dt} log(S(t)) \\
&= - \frac{d}{dt} \left( -\frac{\alpha}{\beta} (e^{\beta t} - 1) \right) \\
&= \frac{d}{dt} \left( \frac{\alpha}{\beta} (e^{\beta t} - 1) \right) \\
&= \frac{d}{dt} \left( \frac{\alpha}{\beta} e^{\beta t} \right) \\
&= \alpha e^{\beta t}
\end{aligned}
$$
$$
f(t) = \alpha \exp(\beta t -\frac{\alpha}{\beta} (e^{\beta t} - 1))
$$

Showing $f(t)$

$$
\lambda(t) = \frac{f(t)}{S(t)} \\
f(t) = \lambda(t) S(t)
$$
$$
\begin{aligned}
f(t) &= \lambda(t) S(t) \\
&= \alpha e^{\beta t} * \exp(-\frac{\alpha}{\beta} (e^{\beta t} - 1)) \\
&= \alpha \exp(\beta t-\frac{\alpha}{\beta} (e^{\beta t} - 1))
\end{aligned}
$$

## b

Modal Time of Death (mode of $f(t)$)

$$
\begin{aligned}
\frac{d}{dt} f(t) &= f(t) * (\beta - \alpha e^{\beta t}) = 0 \\
&\implies (\beta - \alpha e^{\beta t}) = 0 \text{ or } f(t) = 0
\end{aligned}
$$

so the mode is at:

$$ 
t = \frac{log(\frac{\beta}{\alpha})}{\beta}
$$

as long as $\alpha < \beta$

otherwise the function is decreasing so:
$$
t = 0
$$

## c

$ h(x) = ae^{bx} = e^{log(a) + bx}$
So:
$log(h(x)) = log(a) + bx$

```{r Fix the data}
df_c <- df1 %>%
	filter(between(age, 40, 100)) %>%
	mutate(loghx = log(hx))

df_1961 <- df_c %>% filter(year == 1961)
df_2011 <- df_c %>% filter(year == 2011)

lm1961 <- lm(loghx ~ age, data = df_1961)
lm2011 <- lm(loghx ~ age, data = df_2011)

coef1961 <- coef(lm1961)
coef2011 <- coef(lm2011)

a1961 = unname(exp(coef1961[1]))
a2011 = unname(exp(coef2011[1]))

b1961 = unname(coef1961[2])
b2011 = unname(coef2011[2])

```

The values for 1961 are: alpha of `r a1961` and beta of `r b1961` compared 
to the values for 2011 : alpha of `r a2011` and beta of `r b2011`

The meaning of alpha is the starting level of mortality (much higher for 1961)
and beta gives the increase in mortality over time which surprisingly is 
higher for 2011. Perhaps lower infant mortality screws with us a tiny bit
and makes it seem like people die faster with age in 2011 than they did in 1961 just
because so many of them already died before 40 where we start.

## d

```{r plotting estimated hazards}

preds_d <- tibble(age = seq(from = 40, to = 100, by = 1))
preds_d$predicted_log_hx_1961 <- predict(lm1961, newdata = preds_d)
preds_d$predicted_log_hx_2011 <- predict(lm2011, newdata = preds_d)
preds_d$actual_log_hx_1961 <- df_1961$loghx
preds_d$actual_log_hx_2011 <- df_2011$loghx

preds_d_long <- preds_d %>% pivot_longer(cols = c("predicted_log_hx_1961", "predicted_log_hx_2011", "actual_log_hx_1961", "actual_log_hx_2011"), names_to = "type", values_to = "log_hazard")
 
preds_d_long %>%
	ggplot(aes(x = age, y = log_hazard, color = type)) +
	geom_point() +
	theme_minimal()

```

They both seem to fit surprisingly well. There are some minor patterns in the
predicted vs actual for 2011 between the ages of 70 and 80 where the 
actual log hazard seems to be lower, and later on when actual log hazard
seems to be higher for those 90+. For 1961 model it seems to be the opposite for
the super old - we overestimate the log hazard for those pushing 100. Overall
I would say the assumption is quite reasonable. 

## e

```{r modal age at death}

mode1961 <- log(b1961 / a1961) / b1961
mode2011 <- log(b2011 / a2011) / b2011

df_e <- df_c %>% filter(year %in% c(2011, 1961))

df_e %>% filter(year == 1961) %>%
	ggplot(aes(x = age, y = dx)) +
	geom_point() +
	geom_vline(xintercept = mode1961, color = "red") +
	theme_minimal() +
	labs(title = "Density for 1961")

df_e %>% filter(year == 2011) %>%
	ggplot(aes(x = age, y = dx)) +
	geom_point() +
	geom_vline(xintercept = mode2011, color = "red") +
	theme_minimal() +
	labs(title = "Density for 2011")

```

## f

I could probably figure this out with map(lm) but I don't wanna right now.

```{r loopin over, echo=TRUE}

years <- unique(df_c$year)
alphas <- rep(NA, length(years))
betas <- rep(NA, length(years))


for (i in 1:length(years)) {
	# print(i)
	 # fit model
	df_model <- df_c %>%
		filter(year == years[i])
	lm_loop <- lm(loghx ~ age, data = df_model)
	coef_model <- coef(lm_loop)

	alphas[i] <- unname(exp(coef_model[1]))
	betas[i] <- unname(coef_model[2])
	
	df_model$loghx_pred <- predict(lm_loop, newdata = df_model)
	
	# df_plot <- df_model %>% pivot_longer(cols = c("loghx_pred", "loghx"), names_to = "type", values_to = "log_hazard")
	# p <- df_plot %>% ggplot(aes(x = age, y = log_hazard, color = type)) +
	# geom_point() +
	# theme_minimal() +
	# labs(title = paste0("log hazards for predicted vs actual for the year ", years[i]))
	# print(p)
}

```
```{r Plot of alphas and betas}

df_all_years <- tibble(year = years, alpha = alphas,beta = betas) %>%
	mutate(mode = log(beta / alpha) / beta)

df_all_years %>%
	ggplot(aes(x = year, y = alpha)) +
	geom_point() +
	theme_minimal() +
	labs(title = "Alphas over time")

df_all_years %>%
	ggplot(aes(x = year, y = beta)) +
	geom_point() +
	theme_minimal() +
	labs(title = "Betas over time")

df_all_years %>%
	ggplot(aes(x = year, y = mode)) +
	geom_point() +
	theme_minimal() +
	labs(title = "Mode over time")

```

TODO: describe

# Question 2

```{r Load data}

deaths <- read_rds(here("data", "infant.RDS"))
births <- read_rds(here("data", "births.RDS"))

```


## a

```{r Q2a}

tot_deaths <- deaths %>% group_by(race) %>% summarize(deaths = n())
tot_births <- births %>% group_by(race) %>% summarize(births = sum(births))

summary_tbl <- left_join(tot_deaths, tot_births) %>%
	mutate(IMR = deaths / births)

kableExtra::kable(summary_tbl, format = "html")

mortalities <- summary_tbl %>% pull(IMR)
ratio <- mortalities[1] / mortalities[2]
```

So we have that the ratio of black to white mortality is `r ratio`

## b

```{r Q2b}
library(survival)

deaths_sum <- deaths %>% group_by(race, prematurity, aged) %>% summarize(deaths = n()) 

km_df <- deaths_sum %>% mutate(deaths_tot = cumsum(deaths)) %>% 
	left_join(births) %>%
	mutate(exposure = births - deaths_tot + deaths,
		   hazard = deaths / exposure, 
		   survival = cumprod(1 - hazard),
		   var_comp = hazard / (exposure - deaths),
		   variance  = survival^2 * cumsum(var_comp),
		   survival_plus_two = survival + 2 * sqrt(variance),
		   survival_minus_two = survival - 2 * sqrt(variance)
	)

```

## c

```{r Q2c}
 
km_df %>% filter(race == "NHB") %>%
	ggplot() +
	aes(x = aged, y = survival, m = survival_plus_two, n = survival_minus_two,  facet = prematurity) +
	geom_path() +
	facet_grid(prematurity~., scales = "free_y") +
	geom_line(aes(y = survival_plus_two), color = "red") +
	geom_line(aes(y = survival_minus_two),color = "red") +
	theme_bw() +
	labs(title = "NHB")

km_df %>% filter(race != "NHB") %>%
	ggplot() +
	aes(x = aged, y = survival, m = survival_plus_two, n = survival_minus_two,  facet = prematurity) +
	geom_path() +
	facet_grid(prematurity~., scales = "free_y") +
	geom_line(aes(y = survival_plus_two), color = "red") +
	geom_line(aes(y = survival_minus_two),color = "red") +
	theme_bw() +
	labs(title = "NHW")
	
```


## d

## e

Because the births are not equal so higher mortality is not the same as high
number of deaths. 

## f

```{r piecewise constant}

times <- c(1, 7, 14, 28, 60, 90, 120)


```


## g